A mechanization of semi-simplicial and semi-cubical sets in the calculus of inductive constructions

### 2022-11-01 paper/paper.tex
paper: polish
paper: squelch a warning about overfull hbox
paper: more detail in diagram

% The unit type
\newcommand{\unittype}{\ensuremath{\mathsf{unit}}}
\newcommand{\unitpoint}{\ensuremath{\star}}

@@

% These are renewed in tables, but let's define them now for in-document
% typesetting
\newcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\newcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\newcommandx{\filler}[2][1,2]{\prim{filler}[][#1][#2][][]}
\newcommandx{\restrf}[1][1]{\restr{frame}[][#1][][][][][]}
\newcommandx{\fullframe}[1][1]{\prim{fullframe}[][#1][][][]}

@@

\subsection{Semi-cubical sets}
\begin{definition}[\Cube]
  $\Cube$ is defined similarly to $\DeltaPlus$, except that we take sequences of $0$, $1$ and $\star$, instead of sequences of $0$ and $\kstar$. We use $0$ and $1$ to indicate opposite faces of a cube.

  \begin{align*}
    \obj(\Cube) & := \Nat                                                                                \\
    \hom(p, n)  & := \{l \in \{0, 1, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
    g \circ f   & :=
    \begin{cases}
      f                  & \text{if $g = \{\}$}                                                            \\
      a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = 0$ or $1$}                          \\
      a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$, $1$, or $\kstar$} \\
    \end{cases} \\
    \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
  \end{align*}

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the straightforward generalization of augmented semi-simplicial sets and semi-cubical sets to an arbirary arity, where morphisms are sequences of elements of an arbirary set $\nu$, extended with $\kstar$. Then, we observe:

@@

\begin{equation*}
  \begin{array}{llll}
    X_0                                & :              & \unittype                                               & \rightarrow \U \\
    X_1                                & :              & \Sigma \kstar: \unittype. \underbrace{\left(
    \begin{array}{c}
        X_0 \kstar
        \\ \times \\
        X_0 \kstar
      \end{array}\right)}_{\layer[1][1]} & \rightarrow \U                                                                            \\
    X_2                                & :              & \Sigma a: (\Sigma \kstar: \unittype. \underbrace{\left(
      \begin{array}{c}
        \Sigma c: \left(
        \begin{array}{c}X_0 \kstar
            \\ \times \\
            X_0 \kstar
          \end{array}
        \right). (X_1 (\kstar, c)
        \\ \times \\
        \Sigma c: \left(
        \begin{array}{l}
            X_0 \kstar
            \\ \times \\
            X_0 \kstar
          \end{array}
        \right). (X_1 (\kstar, c)
      \end{array}
      \right)}_{\layer[2][1]}.\underbrace{\left(
      \begin{array}{c}
        X_1 \left(\kstar, \left(
          \begin{array}{c}
              a.\tl.0.\hd.0, \\
              a.\tl.1.\hd.0
            \end{array}
          \right)\right)
        \\ \times \\
        X_1 \left(\kstar, \left(
          \begin{array}{c}
              a.\tl.0.\hd.1, \\
              a.\tl.1.\hd.1
            \end{array}
          \right)\right)
      \end{array}
      \right)}_{\layer[2][2]}
                                       & \rightarrow \U                                                                            \\
    \ldots
  \end{array}
\end{equation*}

Then, $\fullframe[n]$ is a nesting of $n$ $\Sigma$-types, of which each component is called a $\layer[n]$.

@@

Now, the two fillers $\filler[n][p](d)$ and $\filler[n][p](d)$ cannot be any fillers. They have to represent two opposite fillers, and we need a way to express this. The answer is simple. The partial frame has two faces and these two fillers have to respectively fill the left face and the right face of the partial frame. We thus have to rely on operations $\restrf[L]$ and $\restrf[R]$ which extract the sides of a frame, leading to the refined definition:

\begin{equation*}
  \begin{array}{lll}
    \filler[n+1][p](d:\framep[n][p]) & \defeq & \Sigma a:(\filler[n][p](\restrf[L](d)) \times \filler[n][p](\restrf[R](d))) .\, \filler[n+1][p+1](d,a)
  \end{array}
\end{equation*}

@@

  {\scriptsize
    \begin{equation*}
      \begin{array}{l}
        \begin{array}{lll}
          \mbox{\textit{squares}}     \\
          [3mm]\mbox{\textit{lines}}  \\
          [3mm]\mbox{\textit{points}} \\
          [1mm]\end{array}
        \framebox{$\begin{array}{lll}
                         \overbrace{\raisebox{0.05cm}{\framebox{$
                               \begin{array}{ll}
                      \raisebox{0cm}{\hspace{.5cm}{\layer[1][0]}}              & \framebox{$a_{0\unitpoint}$} \\
                      [1mm]\overbrace{\framebox{$a_{00}$}~\framebox{$a_{01}$}} &
                      \raisebox{-0.1cm}{\hspace{0.1cm}\filler[1][0]\!\!\!\!\!\!\!\!}
                      \\
                    \end{array}
                             $}}~
                         {\raisebox{0.05cm}{\framebox{$
                               \begin{array}{ll}
                      \raisebox{0cm}{\hspace{.5cm}{\layer[1][0]}}              & \framebox{$a_{1\unitpoint}$} \\
                      [1mm]\overbrace{\framebox{$a_{10}$}~\framebox{$a_{11}$}} &
                      \raisebox{-0.1cm}{\filler[1][0]\!\!\!\!\!\!\!\!}
                    \end{array}
                             $}}}}^{\layer[2][0]}~\stackrel{\raisebox{9.5mm}{$\framebox{$
                                   \begin{array}{ll}
                          \raisebox{0cm}{\hspace{.5cm}{$\layer[2][1]$}}                                & \framebox{$a_{\unitpoint\unitpoint}$} \\
                          [1mm]\overbrace{\framebox{$a_{\unitpoint 0}$}~\framebox{$a_{\unitpoint 1}$}} &
                          \raisebox{-0.1cm}{ \hspace{0.1cm}\filler[2][1]\!\!\!\!\!\!\!\!}
                        \end{array}$}$}}
                         {\raisebox{-0.4cm}{\hspace{2.5cm}\filler[2][0]\!\!\!\!\!\!\!}}
                       \end{array}$}
        ~~~\parbox{2cm}{additionally, each atomic component at dimension $n$ is a \filler[n][n]} \\\\
        [-1mm]\qquad\qquad\quad~~
        \begin{tikzcd}[column sep=5.2em, row sep=1em]
          \phantom{\bullet}\arrow[r, leftrightarrow] & \phantom{\bullet}
        \end{tikzcd}
        \mbox{\framep[1][1]}
        \qquad
        \begin{tikzcd}[column sep=5.2em, row sep=1em]
          \phantom{\bullet}\arrow[r, leftrightarrow] & \phantom{\bullet}
        \end{tikzcd}
        \mbox{\framep[1][1]}                                                                     \\
        \qquad\qquad~~
        \begin{tikzcd}[column sep=32em, row sep=1em]
          \phantom{\bullet}\arrow[r, leftrightarrow] & \phantom{\bullet}
        \end{tikzcd}
        \mbox{\framep[2][1]}                                                                     \\
        \qquad\qquad~~
        \begin{tikzcd}[column sep=39em, row sep=1em]
          \phantom{\bullet}\arrow[r, leftrightarrow] & \phantom{\bullet}
        \end{tikzcd}
        \mbox{\framep[2][2]}                                                                     \\
      \end{array}
    \end{equation*}
  }

>>>>

% The unit type
\newcommand{\unittype}{\ensuremath{\mathsf{unit}}}
\newcommand{\unitpoint}{\ensuremath{\ast}}

@@

% These are renewed in tables, but let's define them now for in-document
% typesetting
\newcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\newcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\newcommandx{\filler}[2][1,2]{\prim{filler}[][#1][#2][][]}
\newcommandx{\restrf}[3][1,2,3]{\restr{frame}[][#3][][#1][#2][][]}
\newcommandx{\fullframe}[1][1]{\prim{fullframe}[][#1][][][]}

@@

\subsection{Semi-cubical sets}
\begin{definition}[\Cube]
  $\Cube$ is defined similarly to $\DeltaPlus$, except that we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$. We use $L$ and $R$ to indicate opposite faces of a cube.

  \begin{align*}
    \obj(\Cube) & := \Nat                                                                                \\
    \hom(p, n)  & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
    g \circ f   & :=
    \begin{cases}
      f                  & \text{if $g = \{\}$}                                                            \\
      a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
      a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
    \end{cases} \\
    \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
  \end{align*}

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the straightforward generalization of augmented semi-simplicial sets and semi-cubical sets to an arbitrary cardinal, where morphisms are sequences of elements of an arbitrary set $\nu$, extended with $\kstar$. Then, we observe:

@@

\begin{equation*}
  \begin{array}{llll}
    X_0                                :                & \underbrace{\unittype}_{\framep[0][0]}                                                   & \rightarrow \U \\
    X_1                                :                & \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
    \begin{array}{c}
        \underbrace{X_0(\unitpoint)}_{\filler[0][0]}
        \\ \times \\
        \underbrace{X_0(\unitpoint)}_{\filler[0][0]}
      \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} & \rightarrow \U                                                                                            \\
    X_2                                :                & \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
        \begin{array}{c}
          \underbrace{\Sigma b: \left(
          \begin{array}{c}
              X_0(\unitpoint)
              \\ \times \\
              X_0(\unitpoint)
            \end{array}
          \right).\; \underbrace{X_1 (\unitpoint, b)}_{\filler[1][1]}}_{\filler[1][0]}
          \\ \times \\
          \underbrace{\Sigma b: \left(
          \begin{array}{c}
              X_0(\unitpoint)
              \\ \times \\
              X_0(\unitpoint)
            \end{array}
          \right).\; \underbrace{X_1 (\unitpoint, b)}_{\filler[1][1]}}_{\filler[1][0]}
        \end{array}
        \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
      \begin{array}{c}
        \underbrace{X_1 \underbrace{\left(a.\hd, \left(
          \begin{array}{c}
            a.\tl.L.\hd.L, \\
            a.\tl.R.\hd.L
          \end{array}
          \right)\right)}_{\restrf[2][1][L]}}_{\filler[1][1]}
        \\ \times \\
        \underbrace{X_1 \underbrace{\left(a.\hd, \left(
          \begin{array}{c}
            a.\tl.L.\hd.R, \\
            a.\tl.R.\hd.R
          \end{array}
          \right)\right)}_{\restrf[2][1][R]}}_{\filler[1][1]}
      \end{array}
      \right)}_{\layer[2][1]}}_{\framep[2][2]}
                                                        & \rightarrow \U                                                                                            \\
    \ldots
  \end{array}
\end{equation*}

$\fullframe[n]$ is a nesting of $n$ $\Sigma$-types, of which each component is called a $\layer[n]$.

@@

Now, the two fillers $\filler[n][p](d)$ and $\filler[n][p](d)$ cannot be any fillers. They have to represent two opposite fillers, and we need a way to express this. The answer is simple. The partial frame has two faces and these two fillers have to respectively fill the left face and the right face of the partial frame. We thus have to rely on operations $\restrf[n][p][L]$ and $\restrf[n][p][R]$ which extract the sides of a frame, leading to the refined definition:

\begin{equation*}
  \begin{array}{lll}
    \filler[n+1][p](d:\framep[n][p]) & \defeq & \Sigma a:(\filler[n][p](\restrf[n][p][L](d)) \times \filler[n][p](\restrf[n][p][R](d))) .\, \filler[n+1][p+1](d,a)
  \end{array}
\end{equation*}

@@
